---
title: "My Exercise - Data.Tables"
author: "Rockefeller University, Bioinformatics Resource Centre"
date: "https://rockefelleruniversity.github.io/RU_course_template/"
output: 
  html_document:
    number_sections: false  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
params:
  toMessage: false    
---

```{r, setup,include=FALSE}
toMessage <- ifelse(!is.null(params$toMessage),params$toMessage,FALSE)
message(toMessage)
```

<p>&nbsp;</p>
These exercises are about [Data.Tables](https://rockefelleruniversity.github.io/Intermediate_R/). 

**Exercise 2** - **Data.Tables**

- Read in the data at: *data/SE216319_TPM_PerSampleTable.csv*

```{r, eval=F, echo=toMessage}
library(data.table)

TPMs <- fread("data/GSE216319_TPM_PerSampleTable.csv")
TPMs

```

- Extract out the unique symbols
```{r, eval=F, echo=toMessage}
sym_uni <- unique(TPMs[,Symbols])
head(sym_uni)
```

- Create a data.table with just the Symbol and IDs (in that order).

```{r, eval=F, echo=toMessage}
sym2ID <- TPMs[,.(Symbols, IDs)]
head(sym2ID)
```

- Do the same but ensure they are ordered by TPM

```{r, eval=F, echo=toMessage}
TPM_ord <- TPMs[order(-TPMs)]
sym2ID <- TPM_ord[,.(Symbols, IDs)]
head(sym2ID)
```

- Subset the data to look at Gapdh

```{r, eval=F, echo=toMessage}
TPMs[Symbols=="Gapdh"]

```

- Calculate the mean TPM for IECs_Vehicle_Mock_rep1_TPM

```{r, eval=F, echo=toMessage}
TPMs[Sample == "IECs_Vehicle_Mock_rep1_TPM",
               .(mean_TPMs = mean(TPMs))]

```

- Calculate the mean TPM for all samples in one-line
```{r, eval=F, echo=toMessage}
TPMs[,.(mean_TPMs = mean(TPMs)), by =Sample]

```

- Create a column that labels if the TPM is equal to 0.

```{r, eval=F, echo=toMessage}
TPMs[TPMs == 0, flag := TRUE]

```

- Create a key by gene symbol

```{r, eval=F}
setkey(TPMs, Symbols)
TPMs[.("Gapdh")]
```

- Do the same but use an index for ID. Again look for Gapdh (14433)

```{r, eval=F}
setindex(TPMs, IDs)
TPMs[.(14433), on="IDs"]
```
