---
title: "Intermediate R, Session 1"
subtitle: " <html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html> Bioinformatics Resource Center - Rockefeller University"
author: 
  - "http://rockefelleruniversity.github.io/Intermediate_R/"
  - "brc@rockefeller.edu"
output: 
  xaringan::moon_reader:
    css: ["default.css", "metropolisCustom.css", "metropolis-fontsCustom.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      titleSlideClass: [middle, inverse]
  html_document:
    toc: true # table of content true
    toc_float: yes
    depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: false  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
params:
  isSlides: "no"
---

```{r,include=FALSE}
suppressPackageStartupMessages(require(knitr))
knitr::opts_chunk$set(echo = TRUE, tidy = T)
```

```{r, results='asis',include=TRUE,echo=FALSE}
if(params$isSlides != "yes"){
  cat("# Intermediate R

---
"    
  )
  
}

```

## Overview

- [Intro](https://rockefelleruniversity.github.io/Intro_To_R_1Day/r_course/presentations/singlepage/introToR_Session1.html#set-up)
- [Tidyverse](https://rockefelleruniversity.github.io/Intro_To_R_1Day/r_course/presentations/singlepage/introToR_Session1.html#background-to-r)
- [Data.table](https://rockefelleruniversity.github.io/Intro_To_R_1Day/r_course/presentations/singlepage/introToR_Session1.html#data_types_in_r)
- [Parallelization](https://rockefelleruniversity.github.io/Intro_To_R_1Day/r_course/presentations/singlepage/introToR_Session1.html#reading-and-writing-data-in-r)

---
```{r, results='asis',include=TRUE,echo=FALSE}
if(params$isSlides == "yes"){
  cat("class: inverse, center, middle

# Getting better at R

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=720px></html> 

---
"    
  )
}else{
  cat("# Getting better at R

---
"    
  )
  
}

```


## Getting better with R

When people start to run R regularly, often they will hit similar roadblocks.

* Reading/writing code is slow
  - More complex R operations can be inconsistent or unnecessarily complex
  - Code can become nested 

* Running code is slow
  - As data gets bigger, R can hit processing or memory limits that slow everything down. 

---
## How do we deal with this?

* Reading/writing code is slow
  - Tidyverse

* Running code is slow
  - Data.table
  - Parallelization

---
## R ecosystems

There are many different ecosystems or communities that exist within R. A popular example for bioinformatics is Bioconductor. Packages associated with Bioconductor have a shared ethos and approach designed to make life easier for the user. 

Generally when you start using R you are using **Base R**. This is the set of functions and tools that come with into R by default. **Base R** is essentially the “core” of the R language. This includes basic data structures i.e. data frames, basic operations i.e. ==, and basic functions i.e. merge(). Most packages in R will be built on **Base R**, so this is why we focus on it during our [Intro course](https://rockefelleruniversity.github.io/Intro_To_R_1Day/). 

---
## R ecosystems

  .pull-left[
**Base R** is great as...
1. There is no need to install extra packages
2. It is pretty lightweight
3. Gives you quite a lot of control
]

  .pull-right[
  but...
1. Syntax can be verbose or inconsistent.
2. Some tasks can get quite tedious once you hit a little bit of complexity.
3. The above issues can make writing code hard, and reading code even harder.
]

---
## R ecosystems

To deal with some of the drawbacks associated with base R, **Tidyverse** was developed.

**Tidyverse** is a collection of packages such as *dplyr*, *tibble*, *ggplot2*, *stringr*. They were designed around a core set of **Tidy** principles that facilitate easy reading/writing and data analysis.

**Tidyverse** also have great community and strong documentation and is well supported as it is integral to Posit/RStudio. Hadley Wickham (Chief Scientist at Posit/RStudio) is the driving force behind the tidyverse. 

---
## R ecosystems

As good as Tidyverse is, it doesn't handle the speed issues R can face (and in fact can often be slower than base R), so in these cases an alternative system can be useful: **Data.table**.


---
## This is all to much!

At this point you may be wondering why we learn 3 different ways to code in R. 

The reality is most folks use these 3 approaches interchangeable but in a specific targeted manner. 

---
## When did you pick each?

* Use tidyverse for clarity, documentation, and community support.
* Use data.table for scale, speed, and efficient memory use.
* Use base R when you need full control, minimal dependencies, or are working in older environments.






